Fusing Filters with Integer Linear Programming

### 2014-05-08 2014betterfusionforfilters/section/02-CNF.tex
Fix citation in section 2
Minor changes

% -----------------------------------------------------------------------------
\section{Combinator Normal Form}
Each input program is expressed in \emph{Combinator Normal Form} (CNF), which is a textual description of its data flow graph. The grammar for CNF is in Figure~\ref{f:CombinatorNormalForm}. Both the @normalize@ and @normalize2@ examples on the previous page are in CNF, and the matching data flow graph for @normalize2@ is in Figure~\ref{f:normalize2-clusterings}. Our data flow graphs are similar to Loop Communication Graphs (LCGs) from related work in imperative array fusion \CITE. We name edges after the corresponding variable from the CNF form, and edges which are fusion preventing are drawn with a dash through them (as per the edge labeled @sum1@ in Figure~\ref{f:normalize2-clusterings}). In data flow graphs we tend to elide the worker functions to combinators when they are not important to the discussion --- so we don't show the @(+)@ operator on each use of @fold@.

Clusters of operators that are fused into single imperative loops are indicated by dotted lines, and we highlight materialized arrays by drawing them in boxes. In Figure~\ref{f:normalize2-clusterings}, the variables @xs@, @ys1@ and @ys2@ are always in boxes as these are the material input and output arrays of the program. However, in the graph on the far right hand side @gts@ has also been materialized because in this version the producing and consuming operators (@filter@ and @fold@) have not been fused. In Figure~\ref{f:CombinatorNormalForm} note that the bindings have been split into the ones that produce scalar values ($sbind$), and the ones that produce array values ($abind$), and these groupings are represented as open and closed arrow-heads in Figure~\ref{f:normalize2-clusterings}.

@@

The exact form of the worker functions is left unspecified, as it is not important for the discussion. We assume workers are pure, can at least compute arithmetic functions of their scalar arguments, and index into arrays in the environment. We assume that each CNF program considered for fusion is embedded in a larger ``host program'' which handles file IO and the like. Workers are additionally restricted so they can only directly reference the \emph{scalar} variables bound by the local CNF program, though they may reference array variables bound by the host program. All access to locally bound array variables is via the formal parameters of array combinators, which ensures that all data dependencies we need to consider for fusion are explicit in the data flow graph.

>>>>

% -----------------------------------------------------------------------------
\section{Combinator Normal Form}
Each input program is expressed in \emph{Combinator Normal Form} (CNF), which is a textual description of its data flow graph. The grammar for CNF is in Figure~\ref{f:CombinatorNormalForm}. Both the @normalize@ and @normalize2@ examples on the previous page are in CNF, and the matching data flow graph for @normalize2@ is in Figure~\ref{f:normalize2-clusterings}. Our data flow graphs are similar to Loop Communication Graphs (LCGs) from related work in imperative array fusion~\cite{gao1993collective}. We name edges after the corresponding variable from the CNF form, and edges which are fusion preventing are drawn with a dash through them (as per the edge labeled @sum1@ in Figure~\ref{f:normalize2-clusterings}). In data flow graphs we tend to elide the worker functions to combinators when they are not important to the discussion --- so we don't show the @(+)@ operator on each use of @fold@.

Clusters of operators that are fused into single imperative loops are indicated by dotted lines, and we highlight materialized arrays by drawing them in boxes. In Figure~\ref{f:normalize2-clusterings}, the variables @xs@, @ys1@ and @ys2@ are always in boxes as these are the material input and output arrays of the program. However, in the graph on the far right hand side @gts@ has also been materialized because in this version the producing and consuming operators (@filter@ and @fold@) have not been fused. In Figure~\ref{f:CombinatorNormalForm}, note that the bindings have been split those that produce scalar values ($sbind$), and those that produce array values ($abind$), and these groupings are represented as open and closed arrow-heads in Figure~\ref{f:normalize2-clusterings}.

@@

The exact form of the worker functions is left unspecified, as it is not important for the discussion. We assume workers are pure, can at least compute arithmetic functions of their scalar arguments, and index into arrays in the environment. We also assume that each CNF program considered for fusion is embedded in a larger ``host program'' which handles file IO and the like. Workers are additionally restricted so they can only directly reference the \emph{scalar} variables bound by the local CNF program, though they may reference array variables bound by the host program. All access to locally bound array variables is via the formal parameters of array combinators, which ensures that all data dependencies we need to consider for fusion are explicit in the data flow graph.
